---
title: "The most harmful events on population and economy of the United States"
output: html_document
---

## Synopsis

This data analysis answers two questions: across the United States:

1. Which types of events are most harmful with respect to population health?
2. Which types of events have the greatest economic consequences?

To answer on this question I analyse U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database that tracks characteristics of major storms and weather events in the United States between 1950 and November 2011

## Data Analysis

### Prerequisits

```{r, message=FALSE}
# R.utils requires to unzip the data file
library("R.utils")
# ggplot requires to draw graphs
library("ggplot2")
```

### Load NOAA database into R

```{r, cache=TRUE}
# download and unpack data from the Internet
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              destfile = "repdata-data-StormData.csv.bz2")
bunzip2("repdata-data-StormData.csv.bz2", overwrite = TRUE)

# load data into R
data = read.csv("repdata-data-StormData.csv")
```

### Show information over the dataset

```{r, cache=TRUE}
# data dimentions
rows <- dim(data)[1]
columns <- dim(data)[2]

# show all columns
colnames(data)
```

### Remove unused data

A big dataset with `r rows` rows and `r columns` columns is rather big so I keep the columns I need:
```{r, cache=TRUE}
tinyData <- data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
colnames(tinyData) <- c("EventType", "Fatalities", "Injuries", "PropertyDamage", "PropertyDamageMultiplicator", "CropDamage", "CropDamageMultiplicator")
colnames(tinyData)
```

### Change the data type from strings to numbers

The columns `PropertyDamageMultiplicator` and `CropDamageMultiplicator` have the symbols 'H', 'K', 'M' and 'B' instead of numbers 100, 1000, 1000000, 1000000000
```{r}
summary(tinyData$PropertyDamageMultiplicator)
summary(tinyData$CropDamageMultiplicator)
```

and I replace the symbols with the numbers, multiply and remove the column "...Multiplicator"
- for Property Damage
```{r, cache=TRUE}
levels(tinyData$PropertyDamageMultiplicator)[1:4] <- c("0", "1", "1", "1")
levels(tinyData$PropertyDamageMultiplicator)[10:15] <- c("1000000000", "100", "100", "1000", "1000000", "1000000")
levels(tinyData$PropertyDamageMultiplicator)
tinyData$PropertyDamageMultiplicator <- as.integer(as.character(tinyData$PropertyDamageMultiplicator))
tinyData$PropertyDamage <- tinyData$PropertyDamage * tinyData$PropertyDamageMultiplicator
tinyData$PropertyDamageMultiplicator <- NULL
```

- for Crop Damage
```{r, cache=TRUE}
levels(tinyData$CropDamageMultiplicator)[1:2] <- c("0", "1")
levels(tinyData$CropDamageMultiplicator)[4:8] <- c("1000000000", "1000", "1000", "1000000", "1000000")
levels(tinyData$CropDamageMultiplicator)
tinyData$CropDamageMultiplicator <- as.integer(as.character(tinyData$CropDamageMultiplicator))
tinyData$CropDamage <- tinyData$CropDamage * tinyData$CropDamageMultiplicator
tinyData$CropDamageMultiplicator <- NULL
```

### Look for the most harmful event on population health

- get sorted summary of fatalities
```{r, cache=TRUE}
fatalities <- with(tinyData, aggregate(Fatalities, by = list(EventType), sum))
colnames(fatalities) <- c("EventType", "Fatalities")
fatalities <- fatalities[order(-fatalities$Fatalities),]
```

- get sorted summary of injuries
```{r, cache=TRUE}
injuries <- with(tinyData, aggregate(Injuries, by = list(EventType), sum))
colnames(injuries) <- c("EventType", "Injuries")
injuries <- injuries[order(-injuries$Injuries),]
```

- get sorted summary of fatalities and injuries
```{r, cache=TRUE}
casualties <- merge(fatalities, injuries, by.y = "EventType")
casualties$Casualties <- casualties$Fatalities + casualties$Injuries
casualties <- casualties[order(-casualties$Casualties),]
```

### Look for the most harmful event with the greatest economic consequences

- get sorted summary of property damage
```{r, cache=TRUE}
propertyDamage <- with(tinyData, aggregate(PropertyDamage, by = list(EventType), sum))
colnames(propertyDamage) <- c("EventType", "PropertyDamage")
propertyDamage <- propertyDamage[order(-propertyDamage$PropertyDamage),]
```

- get sorted summary of crop damage
```{r, cache=TRUE}
cropDamage <- with(tinyData, aggregate(CropDamage, by = list(EventType), sum))
colnames(cropDamage) <- c("EventType", "CropDamage")
cropDamage <- cropDamage[order(-cropDamage$CropDamage),]
```

- get sorted summary of damage to property and crop
```{r, cache=TRUE}
damage <- merge(propertyDamage, cropDamage, by.y = "EventType")
damage$Damage <- damage$PropertyDamage + damage$CropDamage
damage <- damage[order(-damage$Damage),]
```

## Results

### Impact on polulation

The most harmful event for humans is `r casualties[1,1]` with the total casualties `r casualties[1,4]` (fatalities `r casualties[1,2]` and injuries `r casualties[1,3]`)

### Impact on economy

The most harmful event on economy is `r damage[1,1]` with the total damage `r damage[1,4]` (the property damage `r damage[1,2]` and the crop damage `r damage[1,3]`)

